---
title: "WMATA"
output: html_document
---

---
title: "Line Chart Refresher: WMATA Ridership"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Quick Review: Making a Line Chart

Let's do a quick refresher on line charts using WMATA ridership data, since a line chart will be part of your homework. This is daily bus and rail data from January 2024 through February 2025 (you may recall it from the midterm). We'll aggregate it to weekly totals so the chart is easier to read, then make a line chart.

### Task 1: Load the tidyverse and the data

```{r}
library(tidyverse)
library(lubridate)

wmata <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")

head(wmata)
```

### Task 2: Make weekly totals

Daily data is noisy — weekends and holidays create big dips that make trends hard to see. Let's aggregate to weekly totals using `floor_date()` to group days into weeks, then `summarize()` to add them up.

```{r}
weekly_wmata <- wmata |>
  mutate(week = floor_date(date, "week")) |>
  group_by(week) |>
  summarize(
    weekly_total = sum(total),
    days = n()
  ) |>
  filter(days == 7)  # Keep only complete weeks

head(weekly_wmata)
```

### Task 3: Make a line chart

Create a line chart of weekly ridership over time.

```{r}
weekly_wmata |>
  ggplot(aes(x = week, y = weekly_total)) +
  geom_line(color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Weekly WMATA Ridership (2024-2025)",
    x = "Week",
    y = "Total Riders"
  ) +
  theme_minimal()
```

**Answer this question:** Describe the trend you see. Is ridership going up, down, or staying flat? What are the big dips, and what might explain them? What would you want to know before writing a story about Metro weekly ridership trends?

The ridership is has lots of extreme peaks and troughs, but on average it has stayed the same, with March 2025 having a slightly higher number of riders than January 2024. There are major troughs in December and January of 2025, and looking at the other end of the graph, the same can be said for the same time in 2024. I would want to know if these tourhgs are because of people travelling out of the city to go home for the holidays or if there was some other reason. The peaks come from a gradual increase, not a dramatic change like with the down periods, they probably represent an increase of tourism in the pleasant weather of the fall and summer.
