---
title: "final_project"
output: html_document
---
By Lucy Scothern
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Introduction:
for my final project, I analyzed violent crime trends in the jurisdictions surrounding several Big Ten universities from 2021 through to 2026. The main goal of this project is to determine which Big Ten University out of the sample I have chosen has the most dangerous surrounding area based on violent crime rates per 100,000.

My central hypothesis is thatprince George's County, which surrounds the University of Maryland, has the highest violent crime rate among the jurisdictions included in my analysis. Because PG County is often perceived as having higher crime levels than many of the areas the other Big Ten college communities are located in, I wanted to test whether the data supports that assumption.

My project also examines whether the violent crime rates are increasing, decreasing or remaining stable over time to detect patterns that may cause a change in area safety in the future, across the different jurisdictions.

The data comes from the FBI Crime Data Explorer. Each dataset includes monthly totals for aggravated assault, homicide, rape and robbery. 

The jurisdictions included in this project are:
- Bloomington (Indiana University)
- Champaign (University of Illinois)
- East Lansing (Michigan State)
- Evanston (Northwestern)
- Franklin County (Ohio State)
- New Brunswick (Rutgers)
- Prince George's County (University of Maryland)
- Washentaw County (University of Michigan)

After finding that some areas have larger jurisdictions than others, especially PG County, Franklin County and Washentaw County, raw crime totals would unfairly make these bigger areas appear more dangerous. Calculating per 100,000 residents is necessary for accurate comparison.

Another issue in the dataset is missing data in some of the most recent months. Those values are treated as N/A values so that R handles them correctly.

Loading libraries:

```{r}
library(tidyverse)
library(lubridate)
library(broom)
library(knitr)
```

The first step is loading all of the datasets for each jurisdiction:

```{r}
Bloomington_VC_Data_Sheet1 <- read_csv("Bloomington VC Data - Sheet1.csv", na = c("N/A"))
Champaign_VC_Data1 <- read_csv("Champaign VC Data - Sheet1.csv", na = c("N/A"))
East_Lansing_VC_Data1 <- read_csv("East Lansing VC Data - Sheet1.csv", na = c("N/A"))
Franklin_County_VC_Data1 <- read_csv("Evanston VC Data - Sheet1.csv", na = c("N/A"))
New_Brunswick_VC_Data1 <- read_csv("New Brunswick VC Data - Sheet1.csv", na = c("N/A"))
PG_County_VC_Data1 <- read_csv("PG County VC Data - Sheet1.csv", na = c("N/A"))
Washentaw_County_VC_Data1 <- read_csv("Washentaw County VC Data - Sheet1.csv", na = c("N/A"))

```


Clean and Organize Data: Each dataset includes monthly counts for aggravated assault, homicide, rape and robbery. I cleaned the variable names, converted the month column into a date format, added population estimates and calculated VC rates per 100,000 res.

```{r}
Bloomington <- Bloomington_VC_Data_Sheet1 |>
  rename(
    month = Month,
    aggravated_assault= `Aggravated Assault Offenses`,
    homicide = `Homicide Offenses`,
    rape = `Rape Offenses`,
    robbery = `Robbery Offenses`
  ) |> 
  mutate(
    month = my(month),
    population = 80000,
    total_violent_crime = aggravated_assault + homicide + rape + robbery, rate_per_100k = (total_violent_crime / population) * 100000,
    jurisdiction = "Bloomington"
  )
```










