---
title: "final_project"
output: html_document
---
By Lucy Scothern
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Introduction:
for my final project, I analyzed violent crime trends in the jurisdictions surrounding several Big Ten universities from 2021 through to 2026. The main goal of this project is to determine which Big Ten University out of the sample I have chosen has the most dangerous surrounding area based on violent crime rates per 100,000.

My central hypothesis is thatprince George's County, which surrounds the University of Maryland, has the highest violent crime rate among the jurisdictions included in my analysis. Because PG County is often perceived as having higher crime levels than many of the areas the other Big Ten college communities are located in, I wanted to test whether the data supports that assumption.

My project also examines whether the violent crime rates are increasing, decreasing or remaining stable over time to detect patterns that may cause a change in area safety in the future, across the different jurisdictions.

The data comes from the FBI Crime Data Explorer. Each dataset includes monthly totals for aggravated assault, homicide, rape and robbery. 

The jurisdictions included in this project are:
- Bloomington (Indiana University)
- Champaign (University of Illinois)
- East Lansing (Michigan State)
- Evanston (Northwestern)
- Franklin County (Ohio State)
- New Brunswick (Rutgers)
- Prince George's County (University of Maryland)
- Washentaw County (University of Michigan)

After finding that some areas have larger jurisdictions than others, especially PG County, Franklin County and Washentaw County, raw crime totals would unfairly make these bigger areas appear more dangerous. Calculating per 100,000 residents is necessary for accurate comparison.

Another issue in the dataset is missing data in some of the most recent months. Those values are treated as N/A values so that R handles them correctly.

Loading libraries:

```{r}
library(tidyverse)
library(lubridate)
library(broom)
library(knitr)
```

The first step is loading all of the datasets for each jurisdiction:

```{r}
Bloomington_VC_Data_Sheet1 <- read_csv("Bloomington VC Data - Sheet1.csv", na = c("N/A"))
Champaign_VC_Data1 <- read_csv("Champaign VC Data - Sheet1.csv", na = c("N/A"))
East_Lansing_VC_Data1 <- read_csv("East Lansing VC Data - Sheet1.csv", na = c("N/A"))
Evanston_VC_Data1 <- read_csv("Evanston VC Data - Sheet1.csv", na= c("N/A"))
Franklin_County_VC_Data1 <- read_csv("Evanston VC Data - Sheet1.csv", na = c("N/A"))
New_Brunswick_VC_Data1 <- read_csv("New Brunswick VC Data - Sheet1.csv", na = c("N/A"))
PG_County_VC_Data1 <- read_csv("PG County VC Data - Sheet1.csv", na = c("N/A"))
Washentaw_County_VC_Data1 <- read_csv("Washentaw County VC Data - Sheet1.csv", na = c("N/A"))

```


Clean and Organize Data: Each dataset includes monthly counts for aggravated assault, homicide, rape and robbery. I cleaned the variable names, converted the month column into a date format, added population estimates and calculated VC rates per 100,000 res.

```{r}
Bloomington <- Bloomington_VC_Data_Sheet1 |>
  rename(
    month = Month,
    aggravated_assault= `Aggravated Assault Offenses`,
    homicide = `Homicide Offenses`,
    rape = `Rape Offenses`,
    robbery = `Robbery Offenses`
  ) |> 
  mutate(
    month = my(month),
    population = 80000,
    total_violent_crime = aggravated_assault + homicide + rape + robbery, rate_per_100k = (total_violent_crime / population) * 100000,
    jurisdiction = "Bloomington"
  )
```

```{r}
Champaign <- Champaign_VC_Data1 |>
  rename(
    month = Month,
    aggravated_assault= `Aggravated Assault Offenses`,
    homicide = `Homicide Offenses`,
    rape = `Rape Offenses`,
    robbery = `Robbery Offenses`
  ) |> 
  mutate(
    month = my(month),
    population = 80000,
    total_violent_crime = aggravated_assault + homicide + rape + robbery, rate_per_100k = (total_violent_crime / population) * 100000,
    jurisdiction = "Champaign"
  )
```

```{r}
East_Lansing <- East_Lansing_VC_Data1 |>
  rename(
    month = Month,
    aggravated_assault= `Aggravated Assault Offenses`,
    homicide = `Homicide Offenses`,
    rape = `Rape Offenses`,
    robbery = `Robbery Offenses`
  ) |> 
  mutate(
    month = my(month),
    population = 80000,
    total_violent_crime = aggravated_assault + homicide + rape + robbery, rate_per_100k = (total_violent_crime / population) * 100000,
    jurisdiction = "East Lansing"
  )
```

```{r}
Evanston <- Evanston_VC_Data1 |>
  rename(
    month = Month,
    aggravated_assault = `Aggravated Assault Offenses`,
    homicide = `Homicide Offenses`,
    rape = `Rape Offenses`,
    robbery = `Robbery Offenses`
  ) |> mutate(
    month = my(month),
    population = 77000,
    total_violent_crime = aggravated_assault + homicide + rape + robbery,
    rate_per_100k = (total_violent_crime / population) * 100000,
    jurisdiction = "Evanston"
  )
```

```{r}
Franklin_County <- Franklin_County_VC_Data1 |>
  rename(
    month = Month,
    aggravated_assault = `Aggravated Assault Offenses`,
    homicide = `Homicide Offenses`,
    rape = `Rape Offenses`,
    robbery = `Robbery Offenses`
  ) |> mutate(
    month = my(month),
    population = 77000,
    total_violent_crime = aggravated_assault + homicide + rape + robbery,
    rate_per_100k = (total_violent_crime / population) * 100000,
    jurisdiction = "Franklin County"
  )
```

```{r}
New_Brunswick <- New_Brunswick_VC_Data1 |>
  rename(
    month = Month,
    aggravated_assault = `Aggravated Assault Offenses`,
    homicide = `Homicide Offenses`,
    rape = `Rape Offenses`,
    robbery = `Robbery Offenses`
  ) |> mutate(
    month = my(month),
    population = 77000,
    total_violent_crime = aggravated_assault + homicide + rape + robbery,
    rate_per_100k = (total_violent_crime / population) * 100000,
    jurisdiction = "New Brunswick"
  )
```

```{r}
PG_County <- PG_County_VC_Data1 |>
  rename(
    month = Month,
    aggravated_assault = `Aggravated Assault Offenses`,
    homicide = `Homicide Offenses`,
    rape = `Rape Offenses`,
    robbery = `Robbery Offenses`
  ) |> mutate(
    month = my(month),
    population = 77000,
    total_violent_crime = aggravated_assault + homicide + rape + robbery,
    rate_per_100k = (total_violent_crime / population) * 100000,
    jurisdiction = "Prince George's County"
  )
```

```{r}
Washtenaw_County <- Washentaw_County_VC_Data1 |>
  rename(
    month = Month,
    aggravated_assault = `Aggravated Assault Offenses`,
    homicide = `Homicide Offenses`,
    rape = `Rape Offenses`,
    robbery = `Robbery Offenses`
  ) |> mutate(
    month = my(month),
    population = 77000,
    total_violent_crime = aggravated_assault + homicide + rape + robbery,
    rate_per_100k = (total_violent_crime / population) * 100000,
    jurisdiction = "Washtenaw County"
  )
```

Combining the data: After cleaning each dataset, I am combining them into one big dataframe using bind_rows(). This makes it possible to compare all of them together instead of analyzing individually.

```{r}
all_crime <- bind_rows(Bloomington,Champaign, East_Lansing, Evanston, Franklin_County, New_Brunswick, PG_County, Washtenaw_County
)
```

Data exploration: Before I started to analyze, I wanted to look at the overall structure of the big dataset and check for missing values or unusual patterns 

```{r}
glimpse(all_crime)
summary(all_crime$rate_per_100k)
```

I can see that the data seems to be consistent across jurisdictions, although some of the most recent months are missing data. The rates vary substantially which means there are meaningful differences worth investigating.

Visualization of VC Over Time: This first visualization is a line chart showing crime rates over time for each jurisdiction. This helps identify any overall trends and any unusual spikes/ declines.

```{r}
all_crime |> ggplot(aes(x= month, y= rate_per_100k, color = jurisdiction)) + geom_line() +
  labs(
    title= "Violent Crime Rates Across Big Ten Jurisdictions",
    x="Month",
    y="Violent crimes per 100,000 residents"
    ) +
  theme_minimal()
```

This initial graph shows substantial differences in VC rates. PG County consistently had the highest violent crime rate throughout most of the time period, often far exceeding the other areas even after adjustment. Most of cluster below 100 violent crimes per 100,000 residents, although Washtenaw County and Champaign show slight spikes on occasion. The chart simply provides early support for my hypothesis.

Next I calculated descriptive stats for each jurisdiction. These include:
- Mean VC rate
- Standard deviation
- Standard error
- Margin of Error 
- 95% confidence intervals

(These measures help estimate a typical violent crime rate & settlign any uncertainty.)

```{r}
crime_summary <- all_crime |>
  group_by(jurisdiction) |> summarize(
    mean_rate = mean(rate_per_100k, na.rm = TRUE),
    sd_rate = sd(rate_per_100k, na.rm = TRUE),
    months = sum(!is.na(rate_per_100k)),
    se_rate = sd_rate / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_rate,
    ci_lower = mean_rate - margin_error,
    ci_upper = mean_rate + margin_error
  )

```
Again these statistics strongly support my hypothesis. PG County's average violent crime rate of approximately 287 crimes per 100,000 residents is much higher than every other jurisdiction.

Washtenaw County & Champaign have the next highest avg rates, but both remain far below PG County. East Lansing and Evanston have the lowest avg violent crime rates in the project.

PG County has largest standard deviation, suggesting that its VC rates tend to fluctuate a lot from month to month compared to many of the other jurisdictions. Also, its confidence interval does not overlap with any of the lower-rate jurisdictions, meaning that its higher crime rate does not come down to random variation in the data.

Visualizing Confidence Intervals:

To achieve a better comparison of the areas, I have created a bar chart showing avg VC rates alongside confidence intervals. CIs give me more context as jurisdictions with large overlap may not be very different from one another whereas jurisdictions with very separated intervals are more likely to have genuinely different crime environments.

```{r}
crime_summary |> ggplot(aes(x = reorder(jurisdiction, mean_rate), y = mean_rate, fill = jurisdiction)) +
  geom_col() + geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  coord_flip() +
  labs(title = "Average Violent Crime Rate by Jurisdiction",
    x = "Jurisdiction",
    y = "Average violent crimes per 100,000 residents") + theme_minimal() + theme(legend.position = "none")
```

- Further support for the hypothesis.
- PG County's avg VC rate is much higher than every other area, and its CI is completely separated from the others. 
- Washtenaw County and Champaign have the next highest avg rates. East Lansing, Evanston and Franklin County have the lowest.

Hypothesis Testing:

Next, I wanted to test whether violent crime rates changed over time. To look at this in the fairest way possible, I split the PG County data into 2 periods- 2021-23 and 2024-26. I then used a t-test to determine whether the avg VC rate changed significantly between the two time periods.

Hypotheses: 
- Null: Thereis no differnece in VC rates between the earlier and later periods.
- Alt Hypothesis: VC rates changed between the two periods.

```{r}
pg_periods <- all_crime |> filter(jurisdiction == "Prince George's County") |> mutate(period = if_else(year(month) <= 2023, "2021-2023", "2024-2026"))

t_test_pg <- t.test(rate_per_100k ~ period, data = pg_periods)

t_test_pg

```

Thee results show a statistically significant decrease in VC rates in PG County over time. The avg rate fell from about 315 crimes per 100k residents during 2021-23 to 250 during 2024-26. The p-value is far below 0.05 meaning that the difference is unlikely to be caused by random variation. The CI backs this up also. Looking at these results, although the area around UMD still has the highest average crime rate in the project, it is following a trend of decreasing crime, as seen in the last few years.

Effect Size: Statistical significance by itself does not tell us if the difference s practically meaningful enough for a story to be written. So, to measure the size of the difference, I calculated Cohen's d.

```{r}
period_summary <- pg_periods |> group_by(period) |>
summarize(mean_rate = mean(rate_per_100k, na.rm = TRUE), sd_rate = sd(rate_per_100k, na.rm = TRUE), months = n())

first_period <- period_summary |> filter(period == "2021-2023")
second_period <- period_summary |> filter(period == "2024-2026")

pooled_sd <- sqrt(
((first_period$months - 1) * first_period$sd_rate^2 + (second_period$months - 1) * second_period$sd_rate^2) /(first_period$months + second_period$months - 2))

cohens_d <- (first_period$mean_rate - second_period$mean_rate) / pooled_sd

cohens_d
```

Cohen’s d: approx. 0.96 = this is considered a large effect size. This outcome suggests that the decrease in VC between 2021-23 and 2024-26 was not only statistically significant but also meaningful in practically. In other words, it appears to represent a substantial real life change rather than a small fluctuation in the stats.

Linear Regression Analysis:

The next step was to use linear regression to estimate the change in VC over time. Instead of splitting the data into 2 periods, regression allows me to estimate a continuous monthly trend. Firstly, I ahd to create a numeric month index for each jurisdiction

```{r}
all_crime <- all_crime |>
  arrange(jurisdiction, month) |>
  group_by(jurisdiction) |>
  mutate(month_index = row_number())
```

