---
title: "Trials_errors_lessons"
output: html_document
---


```{r}
library(tidyverse)
library(lubridate)
```


The feedback you gave me from my last project planning submission was that my t-test is neccessary but not sufficient and to add this week's topic of linear regression to the tests that I do on my data. I will also be adding multiple regression my list of tests after we have gone over it this week.

Here is my code applying your feedback from last week.

First I know that regression needs a number for the x-variable not a date like I have in my dataset- This took me a few tries, I asked google what I needed to do and it said i needed to bind all my data together- I dont know where I'm going wrong here 

```{r}
all_crime <- bind_rows(Bloomington_VC_Data_Sheet1, Champaign_VC_Data_Sheet1, East_Lansing_VC_Data_Sheet1, Evanston_VC_Data_Sheet1, Franklin_County_VC_Data_Sheet1, New_Brunswick_VC_Data_Sheet1, PG_County_VC_Data_Sheet1, Washentaw_County_VC_Data_Sheet1)
```


```{r}

all_crime <- all_crime |> arrange(jurisdiction, month) |> group_by(jurisdiction) |> mutate(month_index = row_number())
```

I need to replace my t.test:

```{r}
pg_model <- all_crime |> filter(jurisdiction == "Prince George's County") |> lm(rate_per_100k ~ month_index, data= _)
summary(pg_model)
```




Honestly I'm very lost with this part of the project right now, I feel like I got off to a really good and detailed start last week, noting down tests that I am still going to use for the most part, but I'm struggling a bit now. I recognize this isn't very detailed but I'd rather ask for pointers first!





